Parametricity for Primitive Nested Types and GADTs

### 2020-12-21 free-theorems-nesteds/20-mscs.tex
revising GADT section - in progress

Generalized algebraic datatypes (GADTs) generalize nested types by...

For example, ...

A natural next step in this line of work is to extend our
parametricity results to GADTs.  It was shown in~\cite{jg08} that (the
data of) GADTs can be represented using object-level left Kan
extensions (over discrete categories), and further shown
in~\cite{jp19} that adding a carefully designed left Kan extension
construct to a calculus supporting primitive nested types preserves
the cocontinuity needed for primitive GADTs to have well-defined,
properly functorial interpretations.

This suggests adding GADTs to our calculus by extending
Definition~\ref{def:wftypes} with such object-level constructs, and
extending Figure~\ref{fig:terms} with categorically inspired
constructs to introduce and eliminate terms of these types, along the
same entirely standard lines as we did there for products, coproducts,
and fixpoints.  In this section we outline how our model from
Section~\ref{sec:term-interp} can be extended to a parametric model
when some classes of GADTs are incorporated as primitives in this
way. However, as we argue in~???? below, we suspect this is not
possible in general.

HERE!!

To show when this can be done, as well as to give a sense of what goes
wrong in the general case, we begin by recalling that GADTs are
naturally interpreted as initial algebras of left Kan extensions, as
suggested in discrete~\cite{jg08} and

Functorial semantics in~\cite{jp19}.

introducing a new
type, $\Lan$, to represent left Kan extensions.  Indeed,

To begin with, we add the following type formation rule for $\Lan$
types to our type system.
\[
\AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A}$}
\TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F) \ol A$}
\DisplayProof
\]
Notice that the length of $\ol K$ coincides with that of $\ol A$.
Introducing the new type also requires the following additional clause in the inductive definition of higher-order substitution.
\[
((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G]
  = (\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi :=_{\ol{\beta}} G]}
\]

Before we extend the term calculus with the term introduction and elimination rules for $\Lan$ types,
it is necessary to generalize $\Nat$ types to bind type constructor variables of arbitrary arity.
So, the formation rule for $\Nat$ types will be modified as
\[
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
\DisplayProof
\]
and the term introduction and elimination rules will be
\[
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\]
and
\[
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\]
respectively.
It is necessary, of course, to suitably extend the semantic interpretations of $\Nat$ types and terms.
The type constructor variables bound by $\Nat$ will be interpreted as either functors
or relation transformers.
The set and relational interpretations for $\Nat$ types are defined as follows.
\begin{align*}
\setsem{\Gamma; \emptyset \vdash \Nat^\Phi\,F\,G}\rho
&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol K] 
  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol K]~|~\\ 
&\hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
&\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol K] \} \\
%%%
\relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\ol{\Phi := K}]
\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
&=
  \{(\eta_1,\eta_2) \in \setsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\emptyset
    \vdash \Nat^\Phi \,F\,G} (\pi_2
  \rho)~|~\\
& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_{k}}.\\
& \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
  (\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi := K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
\end{align*}

We can now give the term introduction and elimination rules for $\Lan$, which are
\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
\DisplayProof
\]
and
\[
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
\DisplayProof
\]
respectively.
Notice that, in these rules, we make essential use of the extended $\Nat$ types.

To give the semantics for $\Lan$ types we follow~\cite{jp19}
and use the fundamental idea that these have to be interpreted as left Kan extensions.
We give the definition of left Kan extension and some useful notation in the following definition.

\begin{defi}
Let $F : \set^k \to \set$ and $K : \set^k \to \set^h$.
Then, the left Kan extension of $F$ along $\ol K$ is given by a functor $\Lan_K\,F : \set^h \to \set$
and a natural transformation $\epsilon : F \to (\Lan_K\,F) \circ K$
which are initial among such pairs of functor and natural transformations, i.e.,
such that for every $L : \set^h \to \set$ and $\eta :  F \to L \circ K$
there exists a unique natural transformation $\mu : \Lan_K\,F \to L$
such that $(\mu K) \circ \epsilon = \eta$, as shown in the following diagram.
\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
\ar[ur, bend left, "{\Lan_K\,F}"{name=Lan, description}]
\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
\ar[Rightarrow, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]
Of course, we can similarly define the left Kan extension for functors on $\rel$.
\end{defi}

One might try to restrict the semantic interpretation of $\Lan$ types to a subset of the set produced by the left Kan extension.
We show in the following proposition that this is not possible, if we want such interpretation to include the interpretation of the terms given by the term-introduction rule.

>>>>

As discussed in Section~\ref{sec:intro}, type indices for nested types
can be any types, including, in the case of truly nested types like
that of bushes, types involving the very same nested type that is
being defined. But every data constructor for a nested type must still
have as its return type exactly the instance being defined. For
example, the data constructors for the instance $\mathtt{PTree\; A}$ of the
nested type $\mathtt{PTree}$ are $\mathtt{pleaf :: PTree\; A}$ and
$\mathtt{pnode :: PTree\; (A \times A) \to PTree\; A}$, and
the data constructors for the instance $\mathtt{Bush \;A}$ of the truly
nested type $\mathtt{Bush}$ are $\mathtt{bnil :: Bush\; A}$ and
$\mathtt{bcons :: A \to Bush\; (Bush\; A) \to Bush\; A}$.

Generalized algebraic datatypes (GADTs){\color{red}~\cite{???}} generalize nested
types by relaxing the above restriction to allow the return types of
data constructors to be different instances of the datatype than the
one being defined. For example, the GADT
{\small
\[\begin{array}{l}
\mathtt{data\; Seq \;(A : Set)\;:\;Set\;where}\\
\hspace*{0.4in}\mathtt{sconst\;:\; A \rightarrow List\;A}\\
\hspace*{0.4in}\mathtt{spair\;:\;Seq\;A \rightarrow Seq\;B \rightarrow
  Seq\;(A\times B)}\\
\hspace*{0.4in}\mathtt{sseq\;:\;(Nat \rightarrow Seq\;A) \rightarrow
  Seq\;(Nat \to A)}
\end{array}\]}
\noindent
\!\!has data constructors $\mathtt{spair}$ and $\mathtt{sseq}$ with
return types $\mathtt{Seq\; (A \times B)}$ and $\mathtt{Seq\;(Nat \to
  A)}$. These types are not only at different instances of \verb|Seq|
from the instance $\mathtt{Seq\; A}$ being defined, but also at
different instances from one another. The resulting interdependence of
different instances of GADTs means that they can express more
constraints than ADTs and nested types. For example, the ADT
$\mathtt{List}$ expresses the invariant that all of the data in the
lists it defines is of the same type, while the nested type
$\mathtt{PTree}$ expresses this invariant {\em as well as} the
invariant that all of the lists it defines have lengths that are
powers of 2. The GADT $\mathtt{Seq}$ enforces {\em even more general}
well-formedness conditions for sequences of values that simply cannot
be expressed with ADTs and nested types alone.

GADTs are widely used in modern functional languages, such as Haskell,
as well as in proof assistants, such as Agda, that are based on
dependent type theories. A natural next step in the line of work
reported in this paper is therefore to extend our parametricity
results to GADTs. A promising starting point for this endeavor is the
observation from~\cite{jg08} that the data objects of GADTs can be
represented using object-level left Kan extensions over discrete
categories. The more recent results of~\cite{jp19} further show that
adding a carefully designed object-level left Kan extension construct
to a calculus supporting primitive nested types preserves the
cocontinuity needed for primitive GADTs to have well-defined, properly
functorial interpretations. Together this suggests extending the type
system in Definition~\ref{def:wftypes} with such a left Kan extension
construct, and extending the calculus in Figure~\ref{fig:terms} with
corresponding categorically inspired constructs to introduce and
eliminate terms of these richer types. This approach exactly mirrors
the (entirely standard) approach taken above for product, coproduct,
and fixpoint types. In this section we outline this approach and show
how our model from Section~\ref{sec:term-interp} can be extended to a
parametric model when some classes of primitive GADTs are incorporated
in this manner. However, as we argue in{\color{red}~????} below, such
extension is not possible in general.

\subsection{Left Kan Extensions}\label{sec:lke}

We begin by recalling the definition of a left Kan extension and
establishing some useful notation surounding them.

\begin{defi}\label{def:lke}
If $F : \set^k \to \set$ and $\ol{K : \set^k \to \set^h}$ are functors
over $\set$, then the {\em left Kan extension of $F$ along $\ol K$} is
a functor $\mathit{Lan}_{\ol K}\,F : \set^h \to \set$ together with a
natural transformation $\epsilon : F \to (\mathit{Lan}_{\ol K}\,F)
\circ \ol K$
%are initial among such pairs of functor and natural transformations,
%i.e.,
such that, for every functor $L : \set^h \to \set$ and natural
transformation $\eta : F \to L \circ \ol K$, there exists a unique
natural transformation $\mu : \mathit{Lan}_{\ol K}\,F \to L$ such that
$(\mu {\ol K}) \circ \epsilon = \eta$. This is depicted in the
following diagram:

\[
\begin{tikzcd}[row sep = huge]
\set^k
\ar[rr, "{F}"{name=Fa}, ""{name=F, below}]
\ar[rd, "{\ol K}"']
&& \set \\
& \set^h
\ar[Rightarrow, bend right = 10, from=F, "{\epsilon}" description]
\ar[ur, bend left, "{\mathit{Lan}_K\,F}"{name=Lan, description}]
\ar[ur, bend right, "{L}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 90, from=Fa, to=Lr, "{\eta}"]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"]
\end{tikzcd}
\]

\noindent
Replacing $\set$ by $\rel$ everywhere in Definition~\ref{def:lke} we
can similarly define the left Kan extension of $F$ along $\ol K$ for
functors $F$ and $\ol K$ over $\rel$.
\end{defi}

{\color{red}

An alternative presentation of left Kan extensions characterizes them
in terms of an adjunction.

Using this adjunction we can represent the types of the data
constructors for $\mathtt{Seq}$ as:
}

\subsection{Extending the Calculus}

In light of the above, we incorporate GADTs into our calculus by first
adding to Definition~\ref{def:wftypes} the following type formation
rule for $\Lan$-types:

\[
\AXC{$\Gamma; \Phi,\ol\alpha^0 \vdash F$}
\AXC{$\ol{\Gamma;\ol\alpha^0 \vdash K}$}
\AXC{$\ol{\Gamma;\Phi \vdash A}$}
\TIC{$\Gamma;\Phi \vdash (\Lan^{\ol{\alpha^0}}_{\ol K} F) \ol A$}
\DisplayProof
\]

\noindent
Here, $\Lan$ binds the variables in $\ol\alpha$, and these variables
must always have arity $0$. In addition, the vectors $\ol K$ and $\ol
A$ must have the same length.

Intuitively, $\Lan^{\ol{\alpha^0}}_K F$ is a syntactic representation
of the left Kan extension of the functor in the variables in
$\ol\alpha$ denoted by $F$ along the functor in the variables in
$\ol\alpha$ denoted by $K$. Using $\Lan$-types, we can therefore
represent the GADT $\mathtt{Seq}$ in our calculus as {\color{red}
  CHECK!!
 \begin{align*}
\mathtt{Seq}\,\alpha &\df \left( \mu \phi.\lambda \beta.
\beta + \left(\Lan^{\gamma_1,\gamma_2}_{\gamma_1 \times \gamma_2}
(\phi \gamma_1 \times \phi \gamma_2) \right) \beta \right. \\ &\qquad
\left. + \left(\Lan^{\gamma}_{\mathtt{Nat} \to \gamma} (\mathtt{Nat} \to \phi
\gamma) \right) \beta + \phi (\mathtt{Nat} \to \beta) \times
\mathtt{Nat} \right) \alpha
\end{align*}
\noindent
where $\mathtt{Nat} = \mu \alpha. \bone + \alpha$ and $\mathtt{Nat}
\to D$ abbreviates $(\Lan^\beta_{\mathtt{Nat}} \bone)D$.} This
representation corresponds exactly to the rewriting of $\mathit{Seq}$
in Section~\ref{sec:lke}. As explained in Section~IV.D of~\cite{jp19},
the more general construct $\Lan^{\ol{\alpha^0}}_{\ol K} F$ allowing
extensions along vectors of functors makes it possible to represent
GADTs with two or more type arguments that depend on one another. Such
GADTs cannot be represented using just unary $\Lan$-types, i.e,
$\Lan$-types of the form $(\Lan^{\ol{\alpha^0}}_K F)\ol A$.
{\color{red} $\Lan$ vs $\Lan$-type.}

To define the term-formation rules for terms of $\Lan$-types we also
need to generalize $\Nat$-types to bind type constructor variables not
just of arity $0$, but of arbitrary arity. {\color{red} This is
  because...WHY????}  Accordingly, we replace the rule for
$\Nat$-types in Definition~\ref{def:wftypes} with

\[
\AXC{$\Gamma;\Phi \vdash F$}
\AXC{$\Gamma;\Phi  \vdash G$}
\BIC{$\Gamma;\emptyset \vdash \Nat^\Phi F \,G$}
\DisplayProof
\]

Finally, adding $\Lan$-types to the types of our calculus also
requires the extension of Definition~\ref{def:second-order-subst} with
the following new clause:
\[((\Lan^{\ol{\alpha}}_{\ol K}\,F) \ol A)[\phi :=_{\ol{\beta}} G] =
(\Lan^{\ol\alpha}_{\ol K}\,F[\phi :=_{\ol{\beta}} G])\, \ol{A[\phi
    :=_{\ol{\beta}} G]}\]

To also extend our term calculus to accommodate GADTs we must first
give introduction and elimination rules appropriate to our generalized
$\Nat$-types. We therefore replace the tenth and eleventh rules in
Figure~\ref{fig:terms} with

\[
\AXC{$\Gamma; \Phi \vdash F$}
\AXC{$\Gamma; \Phi \vdash G$}
\AXC{$\Gamma; \Phi \,|\, \Delta, x : F \vdash t: G$} 
\TIC{$\Gamma; \emptyset
  \,|\, \Delta \vdash L_\Phi x.t : \Nat^\Phi \,F \,G$}
\DisplayProof
\]
and

\[
\AXC{$\ol{\Gamma;\Phi,\ol\beta \vdash K}$}
\AXC{$\Gamma; \emptyset
  \,|\, \Delta \vdash t : \Nat^{\ol\psi} \,F \,G$}
\AXC{$\Gamma;\Phi \,|\, \Delta \vdash s: F[\overline{\psi :=_{\ol\beta} K}]$}
\TIC{$\Gamma;\Phi\,|\, \Delta \vdash t_{\ol K} s:
  G[\overline{\psi :=_{\ol\beta} K}]$}
\DisplayProof
\]
respectively. Finally, we add to the term calculus in
Figure~\ref{fig:terms} the following introduction and elimination
rules for terms of $\Lan$-types, respectively:

\[
\AXC{$\Gamma; \Phi,\ol{\alpha}\vdash F$}
\AXC{$\ol{\Gamma; \ol{\alpha} \vdash K}$}
\BIC{$\Gamma;\emptyset~|~\Delta \vdash \int_{\ol K,F} : \Nat^{\Phi}\, F\, (\Lan^{\ol\alpha}_{\ol{K}}\,  F) \ol{K}$}
\DisplayProof
\]

\[
\AXC{$\Gamma;\emptyset~|~\Delta \vdash \eta : \Nat^{\Phi,\ol\alpha}\,
  F\;G[\ol{\beta := K}]$}
\UIC{$\Gamma;\emptyset~|~\Delta \vdash \partial^{G, \ol K}_F \eta :
\Nat^{\Phi, \ol\beta}\, (\Lan^{\ol\alpha}_{\ol K}\,F)\ol\beta\, G$}
\DisplayProof
\]

\noindent
Note that these rules do indeed make essential use of our new,
generalized $\Nat$-types.

{\color{red} Say how these rules mirror real LKEs just as for
  products, coproducts, etc.}

\subsection{Extending the Semantics}

It is necessary, of course, to suitably extend the semantic
interpretations of $\Nat$ types and terms.  The type constructor
variables bound by $\Nat$ will be interpreted as either functors or
relation transformers.  The set and relational interpretations for
$\Nat$ types are defined as follows.
\begin{align*}
\setsem{\Gamma; \emptyset \vdash \Nat^\Phi\,F\,G}\rho
&= \{\eta : \lambda \ol{K}. \,\setsem{\Gamma; \Phi \vdash F}\rho[\Phi := \ol K] 
  \Rightarrow \lambda \ol{K}.\,\setsem{\Gamma; \Phi \vdash G}\rho[\Phi := \ol K]~|~\\ 
&\hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_k}.\\
&\hspace{0.4in} (\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
: \setsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
\rightarrow \setsem{\Gamma; \Phi \vdash G}\Eq_{\rho}[\Phi := \ol K] \} \\
%%%
\relsem{\Gamma; \emptyset \vdash \Nat^\Phi \,F\,G}\rho
&= \{\eta : \lambda \ol{K}.\,\relsem{\Gamma; \Phi \vdash F}\rho[\ol{\Phi := K}]
\Rightarrow \lambda \ol{K}. \,\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\phi := K}]\}\\
&=
  \{(\eta_1,\eta_2) \in \setsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G} (\pi_1 \rho) \times \setsem{ 
    \Gamma;\emptyset
    \vdash \Nat^\Phi \,F\,G} (\pi_2
  \rho)~|~\\
& \hspace{0.3in} \forall \ol{K = (K^1, K^2, K^*) : RT_{k}}.\\
& \hspace{0.4in} ((\eta_1)_{\ol{K^1}},(\eta_2)_{\ol{K^2}}) \in
  (\relsem{\Gamma; \Phi \vdash G}\rho[\ol{\Phi := K}])^{\relsem{\Gamma;\Phi\vdash F}\rho[\ol{\Phi := K}]} \}\\  
\end{align*}

We intend to interpret $\Lan$-types as actual left Kan extensions. But
in order to guarantee parametricity we might think we have to cut down
their set interpretations by restricting to a well-behaved collection
of the sets produced by such interpretations, just as we did both in
our original interpretations of $\Nat$-types and the above
interpretations of our generalized $\Nat$-types.

{\color{red} REORDER!!!!}

Along similar lines, we might try to restrict the semantic
interpretation of $\Lan$ types to a subset of the set produced by the
left Kan extension. {\color{red} WHY?????} We show in the following
proposition that this is not possible, if we want such interpretation
to include the interpretation of the terms given by the
term-introduction rule.
